---
title: "Introduction to atpolR"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to atpolR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../inst/REFERENCES.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

oldpar <- par(no.readonly = TRUE)
```

# Introduction
ATPOL grid was developed in late '60s in Institute of Botany at Jagiellonian University in Kraków. The backgrounds and methodology is described in [@zajacZalozeniaMetodyczneAtlasu1978; @zajacAtlasDistributionVascular1978]. The extensive mathematical research and GIS implementation was done by Łukasz Komsta and Marek Verey [@komstaRewizjaMatematycznaSiatki2016; @vereyTeoretycznaAnalizaPraktyczne2017]. Algorithms provided by Komsta on [OpenATPOL](https://atpol.sourceforge.io/) are basis for implementation in **atpolR** package.

# Basic usage

## Prepare sample data based on published ATPOL data

In our example we will take a published distribution map of _Erigeron acris_ L. subsp. _acris_ published in  [@zajacAtlasRozmieszczeniaRoslin2019]. The image was scanned with resolution 150 dpi and georefenced using QGIS.

```{r eriacr, fig.width = 7, fig.align='center', fig.cap = "_Erigeron acris_ L. subsp. _acris_ distribution taken from [@zajacAtlasRozmieszczeniaRoslin2019]"}
par(mar = c(0, 0, 0, 0))
tif <- system.file("extdata/eriacr.tif", package = "atpolR")
r <- terra::rast(tif)
terra::plotRGB(r)
```

There is hundreds of records. To get all of them we will use `check_atpol_square()` function, which takes as the arguments the POINT coordinates and a raster, and checks if the value of raster cell corresponding to some arbitrary buffer around the POINT equals zero. As the points are drawn in centers of ATPOL 10 km grid, we will check the values of atpol10k() centroids with an buffer with default radius of 1200 m. Depending of the quality of scan and precision of georeferencing, it might be useful to adjust a buffer a bit.

Our raster usually consist of 3 layers, one for each R, G, B component. The difference between them are visible on Fig. \@ref(fig:rgbraster).

```{r rgbraster, echo = FALSE, message=FALSE, warning=FALSE, results='hide', fig.width = 7, fig.align='center', fig.cap="R, G and B layers of a raster"}
par(mar = c(0, 0, 0, 0))
par(mfrow = c(1, 3))
reds <- colorspace::sequential_hcl(255, "Reds", rev = TRUE)
grns <- colorspace::sequential_hcl(255, "Greens", rev = TRUE)
blus <- colorspace::sequential_hcl(255, "Blues", rev = TRUE)

terra::plot(r[[1]], col = reds,  legend = FALSE, axes = FALSE)
terra::plot(r[[2]], col = grns,  legend = FALSE, axes = FALSE)
terra::plot(r[[3]], col = blus,  legend = FALSE, axes = FALSE)

par(mfrow = c(1, 1))
```

Looking on scan shown on Fig. \@ref(fig:eriacr) we can see a lot of green and blue components. For further analysis we will take the first layer only. Please note: depending of the scan quality and further image process, you may find other layer more useful. 

```{r rraster, echo = FALSE, message=FALSE, warning=FALSE, results='hide', fig.width = 7, fig.asp = 0.7, fig.align='center'}
terra::plot(r[[1]], col = colorspace::sequential_hcl(255, "Reds", rev = TRUE), axes = FALSE)
```

The values of the layer are continuous, from 0 to 255. For simplification and easier analysis we will classify the layer, just assigning two values: 0 and 1. We will create a reclassification matrix `rclmat` and apply `classify()` function form **terra** package.

```{r echo = TRUE, message=FALSE, warning=FALSE, results='hide'}
m <- c(0,120, 0,
       120,255, 1)
rclmat <- matrix(m, ncol=3, byrow=TRUE)
rc <- terra::classify(r[[1]], rclmat, include.lowest = TRUE)
```
```{r reclasified, echo = FALSE, message=FALSE, warning=FALSE, fig.width=7, fig.asp=0.7, fig.align='center', fig.cap="Reclasiffied raster with 0 --- as black and 1 --- as white"}
terra::plot(rc, axes = FALSE, col = c("black", "white"), legend = FALSE)
```

Having the raster prepared we can load the package and apply the `check_atpol_square()` function for all 10k grids:

```{r eriacr-reverse, echo=TRUE, message=FALSE, warning=FALSE}
library(atpolR)

eriacr <- atpol10k()|>
  dplyr::mutate(a = mapply(function(x) check_atpol_square(x, rc), centroid)) |>
  dplyr::filter(a == "YES")
```

Let's display the results:
```{r echo=TRUE, message=FALSE, warning=FALSE}
eriacr
```
There is 709 observations (grids) in data set.

Let's have a closer look on BE square:

```{r BE, echo=TRUE, fig.width=7, fig.asp=0.7, fig.align='center', message=FALSE, warning=FALSE}
BE <- atpol100k() |>
  subset(Name == "BE") |>
  sf::st_bbox()
par(pty = "s")
plot(NA, type = "n", xlim = c(BE[1], BE[3]), ylim = c(BE[2], BE[4]), axes = FALSE, xlab = "", ylab = "")
terra::plot(rc, legend = FALSE, add = TRUE)

atpol100k() |>
  subset(Name == "BE") |>
  sf::st_cast("LINESTRING") |>
  terra::plot(add = TRUE, col = "blue", lwd = 1.2)

eriacr |>
  subset(substr(Name, 1, 2) == "BE") |>
  sf::st_set_geometry("centroid") |>
  terra::plot(pch = 16, cex = 1.2, col = "blue", add = TRUE)
```

So far, so god. Our data set corresponds to published data. In next step we will extend it with our own observations.

Please note, that `check_atpol_square()` function can give false positive results in case of noisy scans. Also it doesn't recognize the different shapes used in original publications for different 

## Extend the data

Let's assume, we want to extend the data set by adding own observations. If we already know the grid name, we can simply filter out the grid created by `atpol10k()` function by the Name. If we don't have the grid, however we have coordinates, we can use `latlon_to_grid()` function. In below example we are using both methods:

```{r myData, echo = TRUE, warning=FALSE}
myData <- atpol10k() |>
  dplyr::filter(Name %in% c("BE68", 
                            latlon_to_grid(51.13619, 16.95069, 4))) |>
  dplyr::mutate(a = "myData")
```

And let's add them to above BE square plot:

```{r echo = TRUE, eval=FALSE}
myData |>
  sf::st_set_geometry("centroid") |>
  terra::plot(pch = 16, cex = 1.8, col = "red", add = TRUE)
```  


```{r echo=FALSE, fig.align='center', fig.asp=0.7, fig.cap="Data set extended with our observations in grids BE48 and BE68", message=FALSE, warning=FALSE, myDataplot, fig.width=7}
BE <- atpol100k() |>
  subset(Name == "BE") |>
  sf::st_bbox()
par(pty = "s")
plot(NA, type = "n", xlim = c(BE[1], BE[3]), ylim = c(BE[2], BE[4]), axes = FALSE, xlab = "", ylab = "")
terra::plot(rc, legend = FALSE, add = TRUE)

atpol100k() |>
  subset(Name == "BE") |>
  sf::st_cast("LINESTRING") |>
  terra::plot(add = TRUE, col = "blue", lwd = 1.2)

eriacr |>
  subset(substr(Name, 1, 2) == "BE") |>
  sf::st_set_geometry("centroid") |>
  terra::plot(pch = 16, cex = 1.2, col = "blue", add = TRUE)

myData |>
  sf::st_set_geometry("centroid") |>
  terra::plot(pch = 16, cex = 1.8, col = "red", add = TRUE)
```  

## Plot it

**atpolR** package provides a function `plot_points_on_atpol()` which can be used to visualize the data set. Let's merge our two data sets (`eriacr` and `myData`) and plot them together. For removing any duplicates we can use `unique.data.frame()` function from base R, or `distinct(Name)` from **dplyr** package.

```{r}
eriacr <- eriacr |>
  rbind(myData) |>
  unique.data.frame()
```

And final plot.

```{r echo = TRUE, eval = FALSE}
plotPoitsOnAtpol(eriacr$centroid, main = "Erigeron acris subsp. acris", cex = 0.6)
```

```{r eriacratpol, echo = FALSE, fig.width=7, fig.asp=1, fig.align='center', fig.cap="Combined dataset drawn on ATPOL grid"}
par(mar = c(0, 0, 0, 0))
plotPoitsOnAtpol(eriacr$centroid, main = "Erigeron acris subsp. acris", cex = 0.6)
```


# Functions description

Two basic functions `latlon_to_grid()` and `grid_to_latlon()` allows to quickly convert geographical coordinates (given in WGS 84 latitude and longitude degrees) to ATPOL grid and from grid to coordinates respectively.

```{r}
latlon_to_grid(51.01234, 17.23456, 4)
latlon_to_grid(51.01234, 17.23456, 6)
```
The firs two arguments `latlon_to_grid()` function are latitude and longitude respectively. The third argument is the length of returned grid; it might be even number between 2 and 12.

```{r}
grid_to_latlon("CE50")
```
By default `grid_to_latlon()` returns the center of grid square. If you wish to get it's corners, you can pass another 2 arguments to the function, which are X and Y offsets, like:
```{r}
grid_to_latlon("CE50", xoffset = 1, yoffset = 1)
```
for bottom right corner.

ATPOL 10km x 10km and 100km x 100k grids are generated by `atpol10k()` and `atpol100k()` functions respectively. It returns set of simple features geometries with grids as polygons:

```{r}
atpol100k()
```

For 10k grid it returns a centroids as well:
```{r}
atpol10k()
```

Please note, that ATPOL grids are projected in EPSG:2180 coordinate reference system, commonly used in Poland. 

Function `atpol_div(grid, divider)` divides any given `grid` to smaller grids by 2, 4 or 5 as proposed in [@vereyStandaryzacjaZapisuPodzialow2018].

```{r atpoldiv, echo = FALSE, message=FALSE, warning=FALSE, results='hide', fig.width = 7, fig.align='center', fig.cap="Division by 2, 4 and 5 with adopted naming convection d, c, p"}
par(mar = c(0, 0, 0, 0))
par(mfrow = c(1, 3))
for (d in c(2, 4, 5)) {
  a <- atpol_div("BE23", d)
  a$centroid <- sf::st_centroid(a$geometry)

  plot(a$geometry)
  a |>
    sf::st_set_geometry("centroid") |>
    subset(select = c("Name", "centroid")) |>
    terra::vect() |>
    terra::text(labels = substr(a$Name,5,7))
  
}
par(mfrow = c(1, 1))
```


```{r boundaryPL, fig.width=6, fig.align='center', fig.cap="Boundary of Poland on ATPOL grid."}
par(mar = c(0, 0, 0, 0))
b <- boundaryPL()
plot(atpol100k()$geometry)
plot(b, col = "red", add = TRUE)
```

```{r, include = FALSE}
par(oldpar)
```


# Bibliography
